Choosing Color Palettes for Statistical Graphics

### 2006-10-23 trunk/vcd/paper/Color/TAS.Rnw
incorporated KH's comments for TAS color paper
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@598 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

Despite these known drawbacks, HSV space is still the predominantly-used color model in
color picker tools \citep{vcd:Moretti+Lyons:2002,vcd:Meier+Spalter+Karelitz:2004}
and the basis for many palettes in statistical software packages. A notable
exception is \pkg{ColorBrewer.org} \citep{vcd:Harrower+Brewer:2003},
an online tool for selecting color palettes for maps, providing sets of colors
for coding various types of information. Here, we take a similar approach
and describe strategies for the choice of color palettes for categorical and numerical
data. Following \cite{vcd:Brewer:1999} and \cite{vcd:Harrower+Brewer:2003}, we distinguish
three types of palettes: qualitative, sequential and diverging. The first is tailored
for coding categorical information and the latter two are aimed at numerical variables.
Unlike \pkg{ColorBrewer.org}, we do not only provide fixed sets of colors (with a limited
number of colors in each set) but describe paths through a perceptually-based color space
so that the user can decide which path exactly should be taken and how many colors should
be selected. The color space employed by us is the Hue-Chroma-Luminance (HCL) color space
\citep[see][]{vcd:Ihaka:2003} and we derive explicit formulae describing the paths through
this space that correspond to different types of palettes. Given a mapping from HCL to RGB
coordinates, our formulas are extremely easy to implement and we provide such an implementation
in the \proglang{R} language \citep{vcd:R:2006} using the powerful \proglang{R} graphics
system \citep[see][]{vcd:Murrell:2006}. A flexible implementation of all palettes described
is provided in the package \pkg{vcd} \citep{vcd:Meyer+Zeileis+Hornik:2006} using the HCL
color implementation from \pkg{colorspace} \citep{vcd:Ihaka:2004}. This gives the user
both a conceptual and computational tool box for experimenting with color palettes for
a particular display.

The remainder of the paper is organized as follows: Section~\ref{sec:spaces} provides
further information on HCL and other color spaces, contrasting in particular the properties
of HSV and HCL colors. Sections~\ref{sec:qualitative}, \ref{sec:sequential}, and \ref{sec:diverging}
introduce HCL-based qualitative, sequential, and diverging palettes, respectively.
Section~\ref{sec:illustrations} shows some brief illustrations for each type of palette
in statistical graphics before Section~\ref{sec:discussion} concludes the paper with a
discussion. Some supplementary \proglang{R} code that facilitates experimentation with
the new palettes is provided in Appendix~\ref{sec:R}.

@@

  png(file = PNG(files[1]), height = 480, width = 720)
  hcl.wheel(luminance = seq(5, 95, length = 6))
  dev.off()
  png(file = PNG(files[2]), height = 480, width = 720)
  hsv.wheel(value = seq(0, 1, length = 6), cone = TRUE)
  dev.off()
}
@
\includegraphics{hsv-cylinder}

@@

A popular implementation of such a color space, available in many graphics and
statistics software packages, are \emph{HSV} (hue, saturation, value) colors.
They are a simple transformation of \emph{RGB} (red, green, blue) colors
and are defined by a triplet $(H, S, V)$ with $H \in [0, 360]$ and $S, V \in [0, 100]$.
HSV space has the shape of a single regular cone (often inflated to a regular
cylinder). Vertical sections through this space are shown in the upper panel
of Figure~\ref{fig:hsv-hcl}, depicting hue and saturation given different value levels.
Although simple to specify and easily available in many computing environments,
HSV colors have a fundamental drawback: its three dimensions map to the three
dimensions of human color perception very poorly. The three dimensions are
confounded which is most easily seen when converting the vertical sections to 
gray scale images in Figure~\ref{fig:hsv-hcl-gray}. Clearly, the brightness of
colors is not uniform over hues and saturations (given value)---therefore,
HSV colors are often not considered to be perceptually based.

To overcome these drawbacks, various color spaces have been suggested
that properly map to the perception dimensions, the most prominent of which are
the CIELUV and CIELAB spaces developed by the \cite{vcd:CIE:2004}. \cite{vcd:Ihaka:2003}
argues that CIELUV colors are typically preferred for use with emissive technologies
such as computer screens which makes them an obvious candidate for implementation
in statistical software packages. By taking polar coordinates in the UV plane of
CIELUV, \emph{HCL} (hue, chroma, luminance) colors are obtained, defined by
a triplet $(H, C, L)$ with $H \in [0, 360]$ and $C, L \in [0, 100]$. HCL space
has the shape of a distorted double cone: the admissable chroma and luminance values
depend on the hue chosen. The lower panel of Figure~\ref{fig:hsv-hcl} shows
vertical sections through this space: each of the resulting hue/chroma planes (given
luminance) is now properly balanced towards the same gray (going from black to white
with increasing luminance) which becomes obvious when converting the colors to
a gray scale is in Figure~\ref{fig:hsv-hcl-gray}. This balancing of HCL colors
gives us the opportunity to conveniently choose color palettes which code categorical
and/or numerical information by translating it to paths along the three perceptual axes.
However, some care is required for dealing with the irregular shape of the HCL space
which will be addressed in the following sections.

All palettes described are 
available in \proglang{R} \citep{vcd:R:2006} in the package \pkg{vcd} \citep{vcd:Meyer+Zeileis+Hornik:2006}
using the HCL color implementation from \pkg{colorspace} \citep{vcd:Ihaka:2004}.
Technical documentation to the \proglang{R} implementations along with a large
collection of examples is available via \verb/help("rainbow_hcl")/ that provides 
more comparisons between existing \proglang{R} palettes (based on HSV colors) and
the HCL color palettes.

\section{Qualitative palettes} \label{sec:qualitative}

@@

In \pkg{vcd}, these palettes are available in the function
\begin{Soutput}
rainbow_hcl(n, c = 50, l = 70, start = 0, end = 360*(n-1)/n, ...)
\end{Soutput}
wher \code{n} controls the number of colors in the palette. The arguments
\code{c} and \code{l} give the fixed chroma and luminance level, respectively,
and \code{start} and \code{end} specify the range of hue angles. The function
is named after the base \proglang{R} function \code{rainbow()} which has a similar
interface but chooses colors in HSV coordinates. It computes the HCL coordinates
and transforms them to RGB coordinates by calling the \code{hcl()} function 
provided by the \pkg{colorspace} package. The \proglang{R} code for reproducing
Figure~\ref{fig:pal-q-examples} is given in the appendix.

\section{Sequential palettes} \label{sec:sequential}

Sequential palettes are used for coding numerical information that simply ranges
in a certain interval where low values are considered to be uninteresting and
high values are interesting. Without loss of generality, we assume that we want
to visualize an intensity or interestingness $i \in [0, 1]$. A typical application
in statistics are heatmaps (see Figure~\ref{fig:bkde}).

The simplest solution to this task is to employ light/dark contrasts, i.e., employ the
most basic perceptual axis. The interestingness is thus coded
by an increasing amount of gray (i.e., decreasing luminance) 
  \[ (H, 0, 90 - i \cdot 60), \]
where the hue $H$ used does not matter, chroma is set to 0 (i.e., no color), and 
luminance ranges in $[30, 90]$ avoiding the extreme colors white ($L = 100$) and 
black ($L = 0$). Instead of going linearly from light to dark gray, luminance
could also be increased nonlinearly, e.g., by some function $f(i)$ that
controls whether intensity/luminance is increased quickly or not. We found $f(i) = i^p$ 
to be a convenient transformation where the power $p$ can be varied to achieve
different degrees of non-linearity.

@@

\setkeys{Gin}{width=.8\textwidth}
\begin{figure}[p]
\begin{center}
<<pal-s-examples, fig=TRUE, height=5, width=6>>=
par(mfrow = c(5, 1), mar = c(0, 0, 0, 0))
pal(sequential_hcl(12, c = 0, power = 2.2))
pal(sequential_hcl(12, power = 2.2))
pal(terrain_hcl(12, c = c(65, 0), l = c(45, 95), power = c(1/3, 1.5)))
pal(heat_hcl(12, c = c(80, 30), l = c(30, 90), power = c(1/5, 1.5)))
pal(rev(heat_hcl(12, h = c(0, -100), l = c(75, 40), c = c(40, 80), power = 1)))
@
\caption{\label{fig:pal-s-examples} Examples for sequential palettes, varying
only luminance (first panel), chroma and luminance (second panel), and hue,
chroma and luminance (remaining panels).}
\end{center}
\end{figure}

In \pkg{vcd}, this strategy is implemented in the function
\begin{Soutput}
sequential_hcl(n, h = 260, c = c(80, 0), l = c(30, 90), power = 1.5, ...) 
\end{Soutput}
where the first element of \code{c} and \code{l} give the starting chroma and luminance
coordinate (by default colorful and dark) and the second element the ending coordinate
(by default light gray). The \code{power} argument implements $p$ and defaults to $1.5$.

To increase the contrast between the colors in the palette even further, the ideas from
the previous sequential palettes can also be combined with qualitative palettes by
simultaneously varying the hue as well:
  \[ (H_2 - i \cdot (H_1 - H_2), C_{\max} - i^{p_1} \cdot (C_{\max} - C_{\min}),
      L_{\max} - i^{p_2} \cdot (L_{\max} - L_{\min})). \]
One application would be a HCL-based version of heat colors that increase from a light yellow
(e.g., $(90, 30, 90)$) to a full red (e.g., $(0, 100, 50)$). To make the change in hue
visible, the chroma needs to increase rather quickly for low values of $i$ and then only
slowly for higher values of $i$. This can be achieved by choosing a power $p_1 < 1$.

In \proglang{R}, these are available in the function
\begin{Soutput}
heat_hcl(n, h = c(0, 90), c = c(100, 30), l = c(50, 90), power = c(1/5, 1), ...)
\end{Soutput}
with which the lower three rows in Figure~\ref{fig:pal-s-examples} are produced. The
\proglang{R} code reproducing this figure and stating the exact parameters used is again
provided in the appendix.

\clearpage

\section{Diverging palettes} \label{sec:diverging}

Diverging palettes are also used for coding numerical information ranging in a certain
interval---however, this interval includes a neutral value. Examples for this include
residuals or correlations (both with the neutral value 0) or binary classification
probabilities (with neutral value 0.5) that could be visualized in mosaic plots (see Figure~\ref{fig:arthritis})
or classification maps (see Figure~\ref{fig:class}). Without loss of generality, we assume
that we want to visualize an intensity or interestingness $i \in [-1, 1]$.

@@

Given sequential palettes, deriving diverging palettes is easy: two different hues are
chosen for adding color to the same amount of `gray' at a given intensity $|i|$.
Figure~\ref{fig:pal-d-construct} shows the chroma/luminance plane back to back for the
hues $H = 0$ and 260 with two different paths---giving slightly different emphasis on
luminance or chroma contrasts---from a full red over a neutral grey to a full blue.
As Figure~\ref{fig:pal-s-construct} illustrates, the pair of hues should be chosen carefully
because the admittable values of in the chrome/luminance plane differ across hues.
Clearly, for deriving symmetric palettes, only colors from the intersection of the
admittable chroma/luminance planes can be used. The particular hues $H = 0$ and $260$
used in Figure~\ref{fig:pal-s-construct} were chosen because they have rather similar
chroma/luminance planes, allowing for both large chroma and luminance contrasts.

In \pkg{vcd}, the following function is provided:
\begin{Soutput}
diverge_hcl(n, h = c(260, 0), c = 80, l = c(30, 90), power = 1.5, ...) 
\end{Soutput}
which has the same arguments as \code{sequential_hcl()} but takes a pair of hues \code{h}.

@@

In this section, we show a collection of examples for the various types of palettes applied
to statistical graphics. The first illustration is a visualization of a bivariate
density estimation for the Old Faithful geyser eruptions data \citep{vcd:Azzalini+Bowman:1990}.
Figure~\ref{fig:bkde}
shows heatmaps of a bivariate kernel density estimate of waiting times between
and duration of geyser eruptions in Yellowstone National Park. Both use a sequential
palette as derived in Section~\ref{sec:sequential} balanced towards the same gray levels
with $L \in [30, 90]$ and $p_2 = 2$. The sequential palette in the left panel uses only gray
colors (i.e., $C_{\max} = 0$) and the palette in the right panel additionally employs
colors with $H \in [0, 90]$, $C \in [30, 80]$ and $p_1 = 1/5$.

\setkeys{Gin}{width=\textwidth}
\begin{figure}[b!]
\begin{center}
<<bkde, height=6, width=12>>=
if(generate_png) {
  png(file = "bkde.png", height = 420, width = 840)
  par(mfrow = c(1, 2))
  data("geyser", package = "MASS")
  dens <- bkde2D(cbind(geyser$duration, geyser$waiting), bandwidth = c(0.2, 3), gridsize = c(201, 201))
  image(dens$x1, dens$x2, sqrt(dens$fhat), xlab = "duration", ylab = "waiting", 
    col = rev(sequential_hcl(33, c = 0, l = c(30, 90), power = c(1/5, 2))))
  box()
  image(dens$x1, dens$x2, sqrt(dens$fhat), xlab = "duration", ylab = "waiting", 
    col = rev(heat_hcl(33, c = c(80, 30), l = c(30, 90), power = c(1/5, 2))))
  box()
  dev.off()
}
@
\includegraphics{bkde}
\caption{\label{fig:bkde} Bivariate density estimation for Old Faithful geyser eruptions.}
\end{center}
\end{figure}

\setkeys{Gin}{width=.6\textwidth}
\begin{figure}[p]
\begin{center}
<<seats, fig=TRUE, height=5, width=6>>=
par(mar = rep(0.8, 4))
seats <- structure(c(226, 61, 54, 51, 222),
  .Names = c("CDU/CSU", "FDP",  "Linke", "Gruene", "SPD"))
parties <- rainbow_hcl(6, c = 60, l = 75)[c(5, 2, 6, 3, 1)]
names(parties) <- names(seats)
pie(seats, clockwise = TRUE, col = parties, radius = 1)
@
\caption{\label{fig:seats} Seats in the German parliament.}
\end{center}
\end{figure}

\setkeys{Gin}{width=.8\textwidth}
\begin{figure}[p]
\begin{center}
<<votes, fig=TRUE, height=9, width=11>>=
data("Bundestag2005")
votes <- Bundestag2005[c(1, 3:5, 9, 11, 13:16, 2, 6:8, 10, 12),
                       c("CDU/CSU", "FDP", "SPD", "Gruene", "Linke")]
parties <- parties[colnames(votes)]
dimnames(votes)[[2]][c(4, 5)] <- c("Gr", "Li")
mosaic(votes, gp = gpar(fill = parties),
  spacing = spacing_highlighting, labeling = labeling_left,
  labeling_args = list(rot_labels = c(0, 90, 0, 0), pos_labels = "center",
  just_labels = c("center","center","center","right"), varnames = FALSE),
  margins = unit(c(2.5, 1, 1, 12), "lines"),
  keep_aspect_ratio = FALSE)
@
\caption{\label{fig:votes} Votes in the German election 2005.}
\end{center}
\end{figure}

To illustrate qualitative palettes, data from the 2005 election for the German parliament `Bundestag'
are employed. In that election, five parties were able to obtain enough votes to enter the
Bundestag---the distribution of seats is depicted in a pie chart in Figure~\ref{fig:seats}.
The colors used are rough metaphors for the political parties, using a red hue $H = 0$ for the
social democrats SPD, a blue hue $H = 240$ for the conservative CDU/CSU, a yellow hue $H = 60$
for the liberal FDP, a green hue $H = 120$ for the green party `Die Gruenen' and a purple 
hue $H = 300$ for the leftist party `Die Linke'. All colors use the same chroma $C = 60$ and
luminance $L = 75$. The pie chart clearly shows that neither the governing coalition of SPD
and Gruene nor the opposition of CDU/CSU and FDP could assemble a majority. Given that no party
would enter a coalition with the leftists, this lead to a big coalition of CDU/CSU and SPD.
Figure~\ref{fig:votes} shows the distribution of votes in this election stratified by 
province (Bundesland) in a highlighted mosaic display. The order of provinces is from 
north to south, first for the 10 Western provinces, then for the 6 Eastern provinces. Clearly,
the SPD performed better in the North and the CDU/CSU better in the South; furthermore,
Die Linke performed particularly well in the Eastern provinces and in Saarland.

As pointed out in Section~\ref{sec:diverging}, diverging palettes are particularly useful
when visualizing residuals or correlations (with natural neutral value 0) or probabilities 
in 2-class supervised learning (with neutral value 0.5). Examples for both situations are
provided here. Figure~\ref{fig:arthritis} visualizes the outcome of a double-blind clinical
trial investigating a new treatment for rheumatoid arthritis. The mosaic rectangles alone
signal that the treatment lead to higher improvement compared to the placebo group which are
shown to be significant by the shading based on the Pearson residuals. Positive residuals, corresponding
to more observations in the corresponding cell than expected under independence, are depicted
in blue, negative residuals in red. Light colors signal significance at 10\% level, full colors
significance at 1\% level. Hence, it can be concluded that there are significantly more
marked improvements in the treated group and significantly fewer in the placebo group than
would be expected under independence between treatment and improvement. More details can
be found in \cite{vcd:Zeileis+Meyer+Hornik:2005}.

Figure~\ref{fig:class} shows the fit of a support vector machine (SVM) to an artificial 
2-class supervised learning example: a mixture of two bivariate normal distributions.
The circles and triangles show the original observations, solid symbols correspond to
the support vectors found. The shading underlying the plot visualizes the fitted decision
values: values around 0 are on the decision boundary and are shaded in light gray, while
regions that are firmly classified to one or the other class are shaded in full blue and
red respectively.

\setkeys{Gin}{width=.7\textwidth}
\begin{figure}[p]
\begin{center}
<<arthritis, fig=TRUE, height=6, width=7>>=
art <- xtabs(~ Treatment + Improved, data = Arthritis, subset = Sex == "Female")
names(dimnames(art))[2] <- "Improvement"
set.seed(1234)
mosaic(art, gp = shading_max)
@
\caption{\label{fig:arthritis} Extended mosaic display for arthritis data.}
\end{center}
\end{figure}

\setkeys{Gin}{width=.7\textwidth}
\begin{figure}[p]
\begin{center}
<<class, height=6, width=8>>=
if(generate_png) {
  png(file = "class.png", height = 480, width = 640)
  set.seed(123)
  x1 <- rmvnorm(75, mean = c(1.5, 1.5), sigma = matrix(c(1, 0.8, 0.8, 1), ncol = 2))
  x2 <- rmvnorm(75, mean = c(-1, -1), sigma = matrix(c(1, -0.3, -0.3, 1), ncol = 2))
  X <- rbind(x1, x2)
  ex1 <- data.frame(class = factor(c(rep("a", 75), rep("b", 75))), x1 = X[,1], x2 = X[,2])
  rm(x1, x2, X)
  fm <- ksvm(class ~ ., data = ex1, C = 0.5)
  plot(fm, data = ex1)
  dev.off()
}
@
\includegraphics{class}
\caption{\label{fig:class} SVM classification plot.}
\end{center}
\end{figure}

>>>>

Despite these known drawbacks, HSV space is still the predominantly-used color model in
color picker tools \citep{vcd:Moretti+Lyons:2002,vcd:Meier+Spalter+Karelitz:2004}
and the basis for many palettes in statistical software packages. A notable
exception is \pkg{ColorBrewer.org} \citep{vcd:Harrower+Brewer:2003},
an online tool for selecting color schemes for maps. It provides a rich
collection of prefabricated palettes (with a fixed maximal number
of colors) and guides its users how to choose a suitable palette for coding
various types of information. Here, we take a similar approach
and describe strategies for the choice of color palettes for categorical and numerical
data. Following \cite{vcd:Brewer:1999} and \cite{vcd:Harrower+Brewer:2003}, we distinguish
three types of palettes: qualitative, sequential and diverging. The first is tailored
for coding categorical information and the latter two are aimed at numerical variables
(or ordinal variables if some numerical coding for its levels is used).
Unlike \pkg{ColorBrewer.org}, we do not only provide fixed sets of colors but suggest
a general principle for selecting colors by traversing paths along perceptual axes in
a suitable color space. Consequently, the user can decide which path exactly should be
taken and how many colors should be selected. The color space employed by us is the
Hue-Chroma-Luminance (HCL) color space \citep[see][]{vcd:Ihaka:2003} and we show how
the general principles suggested can be turned into formulas describing paths of
coordinates in this space. Given a mapping from HCL to RGB coordinates, our formulas are
extremely easy to implement in software and we provide such an implementation
in the \proglang{R} language \citep{vcd:R:2006} using the powerful \proglang{R} graphics
system \citep[see][]{vcd:Murrell:2006}. A flexible implementation of all palettes described
is provided in the package \pkg{vcd} \citep{vcd:Meyer+Zeileis+Hornik:2006} using the HCL
color implementation from \pkg{colorspace} \citep{vcd:Ihaka:2004}. This gives the user
both a conceptual and computational tool box for experimenting with color palettes for
a particular display.

The remainder of the paper is organized as follows: Section~\ref{sec:spaces} provides
further information on HCL and other color spaces, contrasting in particular the properties
of HSV and HCL colors. Section~\ref{sec:palettes} introduces HCL-based qualitative,
sequential, and diverging palettes and Section~\ref{sec:illustrations} shows some brief
illustrations for each type of palette in statistical graphics. Section~\ref{sec:software}
provides some general remarks on the implementation in statistical software packages
as well as some details on our implementation in \proglang{R}. Section~\ref{sec:discussion}
concludes the paper with a discussion.

@@

  png(file = PNG(files[1]), height = 400, width = 600)
  hcl.wheel(luminance = seq(5, 95, length = 6))
  dev.off()
  png(file = PNG(files[2]), height = 400, width = 600)
  hsv.wheel(value = seq(0, 1, length = 6), cone = TRUE)
  dev.off()
}
@
\includegraphics{hsv-cylinder}

@@

A popular implementation of such a color space, available in many graphics and
statistics software packages, are HSV colors. They are a simple transformation of RGB colors
and are defined by a triplet $(H, S, V)$ with $H \in [0, 360]$ and $S, V \in [0, 100]$.
HSV space has the shape of a single regular cone (often inflated to a regular
cylinder). Vertical sections through this space are shown in the upper panel
of Figure~\ref{fig:hsv-hcl}, depicting hue and saturation given different value levels.
Although simple to specify and easily available in many computing environments,
HSV colors have a fundamental drawback: its three dimensions map to the three
dimensions of human color perception very poorly. The three dimensions are
confounded which is most easily seen when converting the vertical sections to 
gray scale images in Figure~\ref{fig:hsv-hcl-gray}. Clearly, the brightness of
colors is not uniform over hues and saturations (given value)---therefore,
HSV colors are often not considered to be perceptually based.

To overcome these drawbacks, various color spaces have been suggested
that properly map to the perception dimensions, the most prominent of which are
the CIELUV and CIELAB spaces developed by the \cite{vcd:CIE:2004}. \cite{vcd:Ihaka:2003}
argues that CIELUV colors are typically preferred for use with emissive technologies
such as computer screens which makes them an obvious candidate for implementation
in statistical software packages. By taking polar coordinates in the UV plane of
CIELUV, HCL colors are obtained, defined by
a triplet $(H, C, L)$ with $H \in [0, 360]$ and $C, L \in [0, 100]$. HCL space
has the shape of a distorted double cone: the admissible combinations of
chroma and luminance coordinates (within the space's boundaries) depend on the hue chosen.
The lower panel of Figure~\ref{fig:hsv-hcl} shows
vertical sections through this space: each of the resulting hue/chroma planes (given
luminance) is now properly balanced towards the same gray (going from black to white
with increasing luminance) which becomes obvious when converting the colors to
a gray scale as in Figure~\ref{fig:hsv-hcl-gray}. This balancing of HCL colors
gives us the opportunity to conveniently choose color palettes which code categorical
and/or numerical information by translating it to paths along the three perceptual axes.
However, some care is required for dealing with the irregular shape of the HCL space
which will be addressed in the following sections.

\section{Color palettes} \label{sec:palettes}

\subsection{Qualitative palettes} \label{sec:qualitative}

@@

\subsection{Sequential palettes} \label{sec:sequential}

Sequential palettes are used for coding numerical information that ranges
in a certain interval where low values are considered to be uninteresting and
high values are interesting. Suppose we need to visualize an intensity or
interestingness $i$ which (without loss of generality) is scaled to the unit
interval. A typical application in statistics are heatmaps (see Figure~\ref{fig:bkde}).

The simplest solution to this task is to employ light/dark contrasts, i.e., rely
on the most basic perceptual axis. The interestingness can be coded
by an increasing amount of gray corresponding to decreasing luminance in HCL space:
  \[ (H, 0, 90 - i \cdot 60), \]
where the hue $H$ used does not matter, chroma is set to 0 (i.e., no color), and 
luminance ranges in $[30, 90]$ avoiding the extreme colors white ($L = 100$) and 
black ($L = 0$). Instead of going linearly from light to dark gray, luminance
could also be increased nonlinearly, e.g., by some function $f(i)$ that
controls whether luminance is increased quickly with intensity or not. We found $f(i) = i^p$ 
to be a convenient transformation where the power $p$ can be varied to achieve
different degrees of non-linearity.

@@

\setkeys{Gin}{width=.8\textwidth}
\begin{figure}[p]
\begin{center}
<<pal-s-examples, fig=TRUE, height=5, width=6>>=
par(mfrow = c(5, 1), mar = c(0, 0, 0, 0))
pal(sequential_hcl(12, c = 0, power = 2.2))
pal(sequential_hcl(12, power = 2.2))
pal(heat_hcl(12, c = c(80, 30), l = c(30, 90), power = c(1/5, 2)))
pal(terrain_hcl(12, c = c(65, 0), l = c(45, 90), power = c(1/2, 1.5)))
pal(rev(heat_hcl(12, h = c(0, -100), l = c(75, 40), c = c(40, 80), power = 1)))
@
\caption{\label{fig:pal-s-examples} Examples for sequential palettes, varying
only luminance (first panel), chroma and luminance (second panel), and hue,
chroma and luminance (remaining panels).}
\end{center}
\end{figure}

To increase the contrast between the colors in the palette even further, the ideas from
the previous sequential palettes can also be combined with qualitative palettes by
simultaneously varying the hue as well:
  \[ (H_2 - i \cdot (H_1 - H_2), C_{\max} - i^{p_1} \cdot (C_{\max} - C_{\min}),
      L_{\max} - i^{p_2} \cdot (L_{\max} - L_{\min})). \]
One application would be an HCL-based version of ``heat colors'' that increase from a light yellow
(e.g., $(90, 30, 90)$) to a full red (e.g., $(0, 100, 50)$). To make the change in hue
visible, the chroma needs to increase rather quickly for low values of $i$ and then only
slowly for higher values of $i$. This can be achieved by choosing a power $p_1 < 1$.
Such a strategy is adopted for the palettes shown in the lower three rows in
Figure~\ref{fig:pal-s-examples} using different pairs of hues as well as different
chroma and luminance contrasts. The exact parameter combinations are provided in the
appendix.

\subsection{Diverging palettes} \label{sec:diverging}

Diverging palettes are also used for coding numerical information ranging in a certain
interval---however, this interval includes a neutral value. Examples for this include
residuals or correlations (both with the neutral value 0) or binary classification
probabilities (with neutral value 0.5) that could be visualized in mosaic plots (see Figure~\ref{fig:arthritis})
or classification maps (see Figure~\ref{fig:class}). Analogously to the previous
section, we suppose that we want to visualize an intensity or interestingness $i$
from the interval $[-1, 1]$ (without loss of generality).

@@

Given useful sequential palettes, deriving diverging palettes is easy: two different hues are
chosen for adding color to the same amount of `gray' at a given intensity $|i|$.
Figure~\ref{fig:pal-d-construct} shows the chroma/luminance plane back to back for the
hues $H = 0$ and 260 with two different paths---giving slightly different emphasis on
luminance or chroma contrasts---from a full red over a neutral grey to a full blue.
As Figure~\ref{fig:pal-s-construct} illustrates, the pair of hues should be chosen carefully
because the admissible values of in the chrome/luminance plane differ across hues.
Clearly, for deriving symmetric palettes, only colors from the intersection of the
admissible chroma/luminance planes can be used. The particular hues $H = 0$ and $260$
used in Figure~\ref{fig:pal-s-construct} were chosen because they correspond to similar
geometric shapes in the chroma/luminance plane, allowing for both large chroma and
luminance contrasts. If potential viewers of the resulting graphic might be color-blind,
the pair of hues should be taken from the yellow/blue axis of the color wheel
rather than the green/red axis as contrasts on the latter axis are more difficult to 
distinguish for color-blind people.

@@

In this section, we show a collection of examples for the various types of palettes applied
to statistical graphics. The first examples illustrates qualitative palettes
and visualizes data from the 2005 election for the German parliament ``Bundestag''
are employed. In this election, five parties were able to obtain enough votes to enter the
Bundestag---the distribution of seats is depicted in a pie chart in Figure~\ref{fig:seats}.
The colors used are rough metaphors for the political parties, using a red hue $H = 0$ for the
social democrats SPD, a blue hue $H = 240$ for the conservative CDU/CSU, a yellow hue $H = 60$
for the liberal FDP, a green hue $H = 120$ for the green party ``Die Gr\"unen'' and a purple 
hue $H = 300$ for the leftist party ``Die Linke''. To obtain rather intense colors, we set
chroma to $C = 60$ and luminance to $L = 75$ for all hues
leading to a much better balanced palette compared to the fully saturated
colors usually found in the (German) media (which instead of blue typically use the extreme
``color'' black for the CDU/CSU). The pie chart clearly shows that neither the governing
coalition of SPD and Gr\"une nor the opposition of CDU/CSU and FDP could assemble a majority.
Given that no party would enter a coalition with the leftists, this lead to a big coalition
of CDU/CSU and SPD. Figure~\ref{fig:votes} shows the distribution of votes in this election
stratified by province (Bundesland) in a highlighted mosaic display. The order of provinces is from 
north to south, first for the 10 western provinces (the former Federal Republic of Germany, FRG),
then for the 6 eastern provinces (the former German Democratic Republic, GDR). 
Clearly, the SPD performed better in the north and the CDU/CSU better in the south; furthermore,
Die Linke performed particularly well in the eastern provinces and in Saarland.

\setkeys{Gin}{width=.6\textwidth}
\begin{figure}[p]
\begin{center}
<<seats, fig=TRUE, height=5, width=6>>=
par(mar = rep(0.8, 4))
seats <- structure(c(226, 61, 54, 51, 222),
  .Names = c("CDU/CSU", "FDP",  "Linke", "Gruene", "SPD"))
parties <- rainbow_hcl(6, c = 60, l = 75)[c(5, 2, 6, 3, 1)]
names(parties) <- names(seats)
names(seats)[4] <- "Grüne"
pie(seats, clockwise = TRUE, col = parties, radius = 1)
@
\caption{\label{fig:seats} Seats in the German parliament.}
\end{center}
\end{figure}

\setkeys{Gin}{width=.8\textwidth}
\begin{figure}[p]
\begin{center}
<<votes, fig=TRUE, height=7, width=8.6>>=
data("Bundestag2005")
votes <- Bundestag2005[c(1, 3:5, 9, 11, 13:16, 2, 6:8, 10, 12),
                       c("CDU/CSU", "FDP", "SPD", "Gruene", "Linke")]
parties <- parties[colnames(votes)]
dimnames(votes)[[2]][c(4, 5)] <- c("Gr", "Li")
mosaic(votes, gp = gpar(fill = parties),
  spacing = spacing_highlighting, labeling = labeling_left,
  labeling_args = list(rot_labels = c(0, 90, 0, 0), pos_labels = "center",
  just_labels = c("center","center","center","right"), varnames = FALSE),
  margins = unit(c(2.5, 1, 1, 12), "lines"),
  keep_aspect_ratio = FALSE)
@
\caption{\label{fig:votes} Votes in the German election 2005.}
\end{center}
\end{figure}

To illustrate sequential palettes, a bivariate density estimation for the Old
Faithful geyser eruptions data \citep{vcd:Azzalini+Bowman:1990} is visualized.
The Old Faithful geyser is one of the most popular sites in Yellowstone National Park
and it is of some interest to understand the relation ship between the duration
of a geyser eruption and the waiting time until the next eruption.
Figure~\ref{fig:bkde} shows (two versions of) a heatmap of a bivariate kernel density estimate of 
duration and waiting time revealing a multi-modal bivariate distribution with more
variation in the waiting time after a longer eruption. Both versions of the heatmap use a sequential
palette as derived in Section~\ref{sec:sequential} balanced towards the same gray levels
with $L \in [30, 90]$ and $p_2 = 1.5$. The sequential palette in the left panel uses only gray
colors (i.e., $C_{\max} = 0$) and the palette in the right panel additionally employs
colors with $H \in [0, 90]$, $C \in [30, 80]$ and $p_1 = 1/5$. The parameters have been 
chosen such that the peaks in the density are brought out very clearly by using a rather
large luminance contrast and $p_2 > 1$.

\setkeys{Gin}{width=\textwidth}
\begin{figure}[t!]
\begin{center}
<<bkde, height=6, width=12>>=
if(generate_png) {
  png(file = "bkde.png", height = 420, width = 840)
  par(mfrow = c(1, 2))
  data("geyser", package = "MASS")
  dens <- bkde2D(cbind(geyser$duration, geyser$waiting), bandwidth = c(0.2, 3), gridsize = c(201, 201))
  image(dens$x1, dens$x2, dens$fhat, xlab = "duration", ylab = "waiting", 
    col = rev(sequential_hcl(33, c = 0, l = c(30, 90), power = c(1/5, 1.5))))
  box()
  image(dens$x1, dens$x2, dens$fhat, xlab = "duration", ylab = "waiting", 
    col = rev(heat_hcl(33, c = c(80, 30), l = c(30, 90), power = c(1/5, 1.5))))
  box()
  dev.off()
}
@
\includegraphics{bkde}
\caption{\label{fig:bkde} Bivariate density estimation for Old Faithful geyser eruptions.}
\end{center}
\end{figure}

As pointed out in Section~\ref{sec:diverging}, diverging palettes are particularly useful
when visualizing residuals or correlations (with natural neutral value 0) or probabilities 
in 2-class supervised learning (with neutral value 0.5). Examples for both situations are
provided here. Figure~\ref{fig:arthritis} visualizes the outcome of a double-blind clinical
trial investigating a new treatment for rheumatoid arthritis. The mosaic rectangles alone
signal that the treatment lead to higher improvement compared to the placebo group; this effect
is shown to be significant by the shading that codes the size of the Pearon residuals.
Positive residuals, corresponding
to more observations in the corresponding cell than expected under independence, are depicted
in blue, negative residuals in red. Light colors signal significance at 10\% level, full colors
significance at 1\% level. Hence, it can be concluded that there are significantly more
marked improvements in the treated group and significantly fewer in the placebo group than
would be expected under independence between treatment and improvement. More details can
be found in \cite{vcd:Zeileis+Meyer+Hornik:2005}.

Figure~\ref{fig:class} shows a heatmap with the fit of a support vector machine (SVM) to an artificial 
2-class supervised learning example: a mixture of two bivariate normal distributions with
different means and covariance matrices.
The circles and triangles show the original observations, solid symbols correspond to
the support vectors found. The shading underlying the plot visualizes the fitted decision
values: values around 0 are on the decision boundary and are shaded in light gray, while
regions that are firmly classified to one or the other class are shaded in full blue and
red respectively.

Both diverging palettes employ a smaller contrast on the luminance axis ($L \in [30, 90]$)
and a larger chroma contrast ($C \in [0, 100]$) compared to the sequential palettes from
the previous example. For Figure~\ref{fig:class}, we set $p_1$ and $p_2$ to the somewhat
smaller value of $1.3$ to bring out all deviations from the decision boundary and not only the
extreme decision values. For Figure~\ref{fig:arthritis}, the choice of
$p_1$ and $p_2$ is not so important because only 5 colors are selected---we use values
$p_1 = p_2 = 1$.
